---
title: "Clusterização Municípios Brasileiros Turismo"
output: html_notebook
---

Clusterização dos Municípios Turísticos do Brasil  
Trabalho de TCC de Graduação em Estatística

```{r}
library(dplyr)
library(tidyr)
library(readxl)
library(janitor)
library(clue)
library(ggplot2)
library(ggridges)
library(patchwork)
library(cluster)
library(factoextra)
library(dbscan)

file_path <- 'data/CLUSTERS_VARIAVEIS_ORIGINAIS_2019.xlsx'
df <- read_excel(file_path)

ibeu_path <- 'data/IBEU_Brasil_Municipios.xlsx'
ibeu_df <- read_excel(ibeu_path)
```


Convert the state abbreviation in df to the state name

```{r}
# Create named vector: abbreviation as names, full state names as values
state_map <- c(
  AC = "Acre",
  AL = "Alagoas",
  AP = "Amapá",
  AM = "Amazonas",
  BA = "Bahia",
  CE = "Ceará",
  DF = "Distrito Federal",
  ES = "Espírito Santo",
  GO = "Goiás",
  MA = "Maranhão",
  MT = "Mato Grosso",
  MS = "Mato Grosso do Sul",
  MG = "Minas Gerais",
  PA = "Pará",
  PB = "Paraíba",
  PR = "Paraná",
  PE = "Pernambuco",
  PI = "Piauí",
  RJ = "Rio de Janeiro",
  RN = "Rio Grande do Norte",
  RS = "Rio Grande do Sul",
  RO = "Rondônia",
  RR = "Roraima",
  SC = "Santa Catarina",
  SP = "São Paulo",
  SE = "Sergipe",
  TO = "Tocantins"
)
df$state_name <- state_map[df$uf]

```

Try to merge the datasets

```{r}
library(stringi)

clean_str <- function(x) {
  x %>%
    tolower() %>%
    stri_trans_general(id = "Latin-ASCII") %>%
    trimws()

}

df <- df %>%
  mutate(across(c(state_name, municipio), clean_str))

ibeu_df <- ibeu_df %>%
  mutate(across(c(UF, Municipio), clean_str))

library(dplyr)
library(stringi)

clean_state_name <- function(x) {
  x %>%
    tolower() %>%
    stri_trans_general(id = "Latin-ASCII") %>%
    gsub("-", "", .) %>%  # remove hyphens like in "espi-rito"
    trimws()
}

ibeu_df <- ibeu_df %>%
  mutate(UF_clean = clean_state_name(UF))


merged_df <- df %>% left_join(ibeu_df, by = c("state_name" = "UF_clean", "municipio" = "Municipio"))
merged_df$IBEU = as.numeric(merged_df$IBEU)
```

Fill missing values: the merge fails for 171 rows, lets fill those NAs with the column-wise average across all non-missing rows

```{r}
target_cols <- c("IBEU", "Mobilidade", "Ambiental", "Habitacional", "Serviços", "Insfraestrutura")
for (col in target_cols) {
  merged_df[[col]] = as.numeric(merged_df[[col]])
}
means <- colMeans(merged_df[target_cols], na.rm = TRUE)
for (col in target_cols) {
  merged_df[[col]][is.na(merged_df[[col]])] <- means[col]
}

```

Scale merged dataset and apply k-means using Manhattan distance

```{r}
continuous_cols <- c(
  "quantidade_empregos",
  "quantidade_estabelecimentos",
  "quantidade_visitas_estimadas_internacional",
  "quantidade_visitas_estimadas_nacional",
  "arrecadacao",
  "IBEU")

df_scale = select(merged_df, continuous_cols)
df_scale[continuous_cols] <- scale(df_scale[continuous_cols])

dist_manhattan <- get_dist(df_scale, method = "manhattan")

set.seed(1)
pam_manhattan <- pam(dist_manhattan, k = 5, nstart = 25)
merged_df$new_cluster <- as.character(pam_manhattan$clustering)

```

Analyse the results: tabulate the original and the new clusters assignment

```{r}
table(merged_df$cluster, merged_df$new_cluster)

```


Analyse the results: describe the distributions of the variables by the new cluster

```{r}
library(dplyr)

cluster_summary <- merged_df %>%
  group_by(new_cluster) %>%
  summarise(
    across(
      c(
        quantidade_empregos,
        quantidade_estabelecimentos,
        quantidade_visitas_estimadas_internacional,
        quantidade_visitas_estimadas_nacional,
        arrecadacao,
        IBEU
      ),
      list(
        mean = ~mean(.x, na.rm = TRUE),
        sd = ~sd(.x, na.rm = TRUE),
        median = ~median(.x, na.rm = TRUE)
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

cluster_summary
```

Let's plot this data using boxplots

```{r}
long_df <- merged_df %>%
  select(new_cluster, all_of(continuous_cols)) %>%
  pivot_longer(-new_cluster, names_to = "variable", values_to = "value")

# Plot
ggplot(long_df, aes(x = factor(new_cluster), y = value)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(x = "Cluster", y = "Value", title = "Boxplots by Cluster") +
  theme_minimal()
```

Save the results in excel format

```{r}
install.packages("openxlsx")
library(openxlsx)
write.xlsx(df, 'data/CLUSTERS_NOVAS_VARIAVEIS_2019.xlsx')
```

